\subsubsection{Classical Optimizer}\label{sec:classical-optimizer}

The \definition{Optimizer} is a component whose goal is to \hl{find the parameters of the quantum circuit that optimize the desired objective function.} It \textbf{never manipulates qubits} because it is a completely classical algorithm running on a classical computer.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{What criteria does the optimizer use to determine the optimal parameters?}}
\end{flushleft}
The optimizer is not guessing randomly. After each execution it \textbf{receives a score}, which is the \textbf{value of} the \textbf{objective function} (e.g., the expectation value of the Hamiltonian) for the current parameters, called also \textbf{cost function}.

\highspace
This means that the \textbf{optimizer is guided by the cost function}, which provides feedback on how good the current parameters are. So the same ansatz (i.e., the same quantum circuit structure) can be used for completely different problems simply by changing the cost function.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{tools} \textbf{How does the optimizer update the parameters?}}
\end{flushleft}
Usually, the optimizer is an \textbf{iterative algorithm} that updates the parameters in each iteration based on the cost function value. The optimization process continues until a stopping criterion is met, such as reaching a maximum number of iterations or achieving a cost function value below a certain threshold. Some common algorithms include gradient-based methods (e.g., gradient descent, Adam, etc.) and gradient-free methods (e.g., Nelder-Mead, COBYLA, etc.).

\highspace
In summary, the Classical \definition{Optimizer} is the classical component of a Variational Quantum Algorithm (VQA). Its role is to find the parameters of the ansatz that optimize the objective function computed from the measurement results. After each execution of the quantum circuit, the optimizer receives the cost value, updates the parameters, and repeats the process. The specific problem is defined by the objective function, while the choice of optimizer is relatively flexible.