\section{Software Defined Networking (SDN)}\label{section: Software Defined Networking (SDN)}

\subsection{Introduction}

\definition{Software-Defined Networking (SDN)} is an \textbf{architectural shift} in networking that \textbf{separates} the \textbf{control plane} from the \textbf{data plane}, allowing for centralized control and programmability. Unlike traditional networks, where control logic is embedded in individual devices, \textbf{SDN introduces a centralized software controller that dynamically manages the entire network}.  

\highspace
The \hl{importance of SDN} lies in its ability to:
\begin{itemize}
    \item Improve \textbf{network flexibility} by enabling real-time changes.
    \item Simplify \textbf{network management} through automation.
    \item Reduce \textbf{hardware dependency} by allowing software-driven policies.
    \item Support \textbf{rapid innovation}, making networks more adaptable.
\end{itemize}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{balance-scale} \textbf{Traditional Networking vs. SDN}}
\end{flushleft}
In \emph{\textbf{traditional networking}}, routers and switches \textbf{contain both}:
\begin{itemize}
    \item \textbf{Control Plane}, decides how traffic should be forwarded (e.g., routing decisions, firewall rules).
    \item \textbf{Data Plane}, physically forwards packets based on the control plane's decisions.
\end{itemize}
Challenges of traditional networking:
\begin{enumerate}[label=\textcolor{Red2}{\alph*.}]
    \item \textcolor{Red2}{\faIcon{times} \textbf{Rigid Configuration}}: Any changes require manual updates on multiple devices.
    \item \textcolor{Red2}{\faIcon{times} \textbf{Vendor Lock-in}}: Hardware manufacturers impose proprietary limitations.
    \item \textcolor{Red2}{\faIcon{times} \textbf{Slow Innovation}}: Implementing new networking features takes years due to hardware constraints.
    \item \textcolor{Red2}{\faIcon{times} \textbf{Complex Management}}: Network engineers must configure each device individually.
\end{enumerate}
How \emph{\textbf{SDN differs}}:
\begin{enumerate}[label=\textcolor{Green3}{\alph*.}]
    \item \textcolor{Green3}{\faIcon{check} \textbf{Control Plane is centralized}} in an SDN controller.
    \item \textcolor{Green3}{\faIcon{check} \textbf{Data Plane remains distributed}} across switches and routers.
    \item \textcolor{Green3}{\faIcon{check} \textbf{Network logic is programmable}}, making updates and changes easier.
\end{enumerate}
This shift makes networks more dynamic, scalable, and easier to manage.