\subsection{Energy Constrained Computing}

\textbf{Energy efficiency is critical} in computing, especially in scenarios where power is limited, such as mobile devices or data centers. Different types of processors consume varying amounts of power based on their energy efficiency and the nature of the tasks they're designed to handle.

\highspace
The following formula shows that \definition{Power Consumption} is the \textbf{product of performance} (operations per second) \textbf{and energy efficiency} (joules per operation):
\begin{equation}
    \text{Power} = \left( \text{Ops}/\text{second} \right) \times \left( \text{Joules}/\text{Op} \right)
\end{equation}
\begin{itemize}
    \item \important{Ops (Operations)}: The number of tasks or operations performed.
    \item \important{Second (Time)}: The duration over which these operations are performed.
    \item \important{Joules (Energy)}: The unit of energy consumed per operation.
    \item $\important{\text{Ops}/\text{second}}$: Operations performed on a second.
    \item $\important{\text{Joules}/\text{Op}}$: Energy used for each operation.
\end{itemize}
It is important to leverage the specialization (using fixed function hardware) to achieve better energy efficiency.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{balance-scale} \textbf{General-Purpose vs Specialized Processors}}
\end{flushleft}
\textbf{Specialized processors are designed for specific tasks, making them more efficient compared to general-purpose processors}.
\begin{itemize}
    \item \definition{General-Purpose Processors (GPPs)}: CPUs are \textbf{designed to handle a wide range of tasks}, making them \textbf{versatile} but \textbf{not always energy-efficient}. They consume more power because they are built to perform various operations reasonably well.

    \item \definition{Specialized Processors (SPs)}: These include GPUs, FPGAs, and custom accelerators \textbf{designed for specific tasks}. Because they focus on particular workloads, they can be \textbf{optimized for energy efficiency}. For example:
    \begin{itemize}
        \item \emph{GPUs}: Excellent for parallel tasks like graphics rendering and matrix operations, consuming less power per operation in those tasks compared to CPUs.
        \item \emph{FPGAs}: Can be tailored to execute specific algorithms with high energy efficiency.
    \end{itemize}
\end{itemize}
