\section*{Preface}

Every theory section in these notes has been taken from the sources:
\begin{itemize}
    \item Validity of the single processor approach to achieving large scale computing capabilities.\cite{amdahl2007validity}
    \item Introduction to parallel algorithms, Carnegie Mellon University.\cite{introductionToParallelAlgorithmsUMD}
    \item Course slides.\cite{parallel-computing-polimi}
    \item Reevaluating Amdahl's law.\cite{gustafson1988reevaluating}
    \item OpenMP by Example, Johnston Hans.\cite{whatIsOpenMPumassJohnstonHans}
    \item Introduction to parallel computing, volume 110.\cite{kumar1994introduction}
    \item NVIDIA Tesla: A unified graphics and computing architecture.\cite{lindholm2008nvidia}
    \item Structured Parallel Programming: Patterns for Efficient Computation.\cite{mccool2012structured}
    \item The critical directive, OpenMP, Microsoft.\cite{openMPCriticalDirectiveMicrosoftExample}
    \item Multithreading Architecture.\cite{nemirovsky2022multithreading}
    \item Cache, Wikipedia.\cite{wikipediaCacheResearch}
\end{itemize}
About:
\begin{itemize}
    \item[\faIcon{github}] \href{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/HPC-E-PoliMI-university-notes}{GitHub repository}
\end{itemize}
These notes are an unofficial resource and shouldn't replace the course material or any other book on parallel computing. It is not made for commercial purposes. I've made the following notes to help me improve my knowledge and maybe it can be helpful for everyone.

As I have highlighted, a student should choose the teacher's material or a book on the topic. These notes can only be a helpful material.