\subsubsection{Lightweight Alternative: Gated Recurrent Unit (GRU)}

\definition{Gated Recurrent Unit (GRU)} were proposed by Cho et al. (2014) as a \textbf{simplified, faster, and lighter} recurrent architecture that retains much of the power of LSTMs but with fewer parameters and a simpler design.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{Why GRU?}} \\
\end{flushleft}
Long Short-Term Memory (LSTM) networks are powerful but can be computationally intensive due to their complex architecture involving multiple gates and memory cells:
\begin{itemize}
    \item[\textcolor{Red2}{\faIcon{times}}] \textbf{Computationally heavy}, because of multiple matrix multiplications per time step.
    \item[\textcolor{Red2}{\faIcon{times}}] \textbf{Contain 3 gates $+$ memory cell}, leading to more parameters to train.
    \item[\textcolor{Red2}{\faIcon{times}}] \textbf{Have many parameters (4 matrix multiplications per time step)}, which can lead to overfitting on smaller datasets.
\end{itemize}
GRUs address these issues by simplifying the architecture while still effectively capturing long-term dependencies in sequential data. Its goal is to:
\begin{itemize}
    \item[\textcolor{Green3}{\faIcon{check}}] \textbf{Reduce computational complexity} by using fewer gates and parameters.
    \item[\textcolor{Green3}{\faIcon{check}}] \textbf{Maintain performance} comparable to LSTMs on many tasks.
    \item[\textcolor{Green3}{\faIcon{check}}] \textbf{Facilitate faster training} and inference times.
    \item[\textcolor{Green3}{\faIcon{check}}] \textbf{Simplify implementation} and tuning due to fewer hyperparameters.
\end{itemize}
They do this by merging some gates and eliminating the explicit memory cell found in LSTMs.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{tools} \textbf{GRU Architecture}}
\end{flushleft}
GRUs \textbf{do not have a separate cell state} $C_t$ like LSTMs. Instead, they only maintain a \textbf{hidden state} $h_t$, which merges: short-term memory and long-term memory. GRUs use \textbf{two gates} to control the flow of information:
\begin{enumerate}
    \item \important{Update Gate $z_t$}. Controls \textbf{how much of the past is kept} and \textbf{how much of the new information is added}.
    \begin{equation}
        \begin{bmatrix}
            i_t \\
            f_t
        \end{bmatrix}
        \quad \xrightarrow{\text{GRU}} \quad
        z_t = \sigma(W_z \cdot [h_{t-1}, x_t] + b_z)
    \end{equation}
    It \textbf{replaces both the input and forget gates} of the LSTM.
    \begin{itemize}
        \item $z_t = 1$: use \textbf{only} \underline{new} information, forget everything from the past (like forget gate closed, $0$, and input gate open, $1$).
        \item $z_t = 0$: keep \textbf{only} \underline{past} information, ignore new input (like forget gate open, $1$, and input gate closed, $0$).
        \item $0 < z_t < 1$: balance between past and new information.
    \end{itemize}


    \item \important{Reset Gate $r_t$}. Controls \textbf{how much of the previous hidden state to consider when creating new candidate} information. This is similar to the input gate in LSTMs, but it directly influences the candidate hidden state. Similarly, it is computed as:
    \begin{equation}
        r_t = \sigma(W_r \cdot [h_{t-1}, x_t] + b_r)
    \end{equation}
    \begin{itemize}
        \item $r_t$ close to 0: ignore past hidden state, focus on new input.
        \item $r_t$ close to 1: consider past hidden state fully, influence of $h_{t-1}$ is strong.
    \end{itemize}
\end{enumerate}
The number of gates is reduced from 3 in LSTMs to 2 in GRUs, \textbf{simplifying dramatically the architecture}. Also, \hl{GRUs do not have a separate memory cell}; instead, they \textbf{maintain a single hidden state that combines both short-term and long-term memory} ($h_t$). This further reduces the number of parameters and computations required.

\highspace
The \textbf{final hidden state} $h_t$ in a GRU is computed as:
\begin{equation}
    h_t = \left(1 - z_t\right) \odot h_{t-1} + z_t \odot \tilde{h}_t
\end{equation}
It interpolates between the previous hidden state, $h_{t-1}$, and the candidate hidden state, $\tilde{h}_t$, weighted by the update gate, $z_t$. Thus, the \textbf{update gate} decides how much past information to retain and how much new information to incorporate. Notably, it \textbf{is the only gate that directly influences the final hidden state}. Instead, the \textbf{reset gate} $r_t$ \textbf{operates when computing the candidate hidden state} $\tilde{h}_t$:
\begin{equation}
    \tilde{h}_t = \tanh(W_h \cdot [r_t \odot h_{t-1}, x_t] + b_h)
\end{equation}

\begin{table}[!htp]
    \centering
    \begin{tabular}{@{} l p{11em} p{11em} @{}}
        \toprule
        \textbf{Feature} & \textbf{LSTM} & \textbf{GRU} \\
        \midrule
        Gates & 3 (input, forget, output) & 2 (update, reset) \\[.5em]
        Memory Cell & Yes ($C_t$) & No (only hidden state $h_t$) \\[.5em]
        Hidden State & Separate from cell state & Combines short-term and long-term memory \\[.5em]
        Complexity & Higher & Lower \\[.5em]
        Parameters & More (4 matrix multiplications per time step) & Fewer (3 matrix multiplications per time step) \\[.5em]
        Training Speed & Slower & Faster \\[.5em]
        Performance & Strong on long sequences & Usually comparable to an LSTM, sometimes better on smaller datasets \\
        \bottomrule
    \end{tabular}
    \caption{Comparison between LSTM and GRU architectures.}
\end{table}

\newpage

\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{Final thoughts: Is GRU always better than LSTM?}}
\end{flushleft}
In short, \textbf{no}, GRU is not the best RNN. In some scenarios, LSTMs may outperform GRUs. They \textbf{do not have the same expressive power} of LSTMs.
\begin{itemize}
    \item \important{LSTMs Separate Long-Term and Short-Term Memory}. LSTM has two dedicated components (\textbf{cell state} $C_t$ and \textbf{hidden state} $h_t$) to manage long-term and short-term dependencies separately. Instead, GRU merges them into a single vector. The separation in LSTMs allows to:
    \begin{itemize}
        \item Preserve long-term memory in $C_t$ \textbf{without exposing it}.
        \item Use $h_t$ for short-term, noisy, rapidly-changing representations.
        \item Control what gets exposed to the next layer (via the output gate).
    \end{itemize}
    This separation is \textbf{very powerful}, especially for tasks requiring deep context management, such as language modeling or machine translation.


    \item \important{LSTMs Have More Control (More Degrees of Freedom)}. LSTM has \textbf{three gates} (forget $f_t$, input $i_t$, output $o_t$), while GRU has \textbf{two} (update $z_t$, reset $r_t$). This gives LSTMs: more expressiveness, finer control of memory flow, and the ability to decouple ``writing to memory'' from ``exposing memory''. In some tasks, this extra structure provides: better accuracy, more stable long-term learning, and better handling of irregular or sparse signals. In summary, \textbf{with LSTM we can do a more nuanced manipulation of information}.


    \item \important{LSTMs Work Better on Very Long Sequences}. Empirically, on tasks requiring \textbf{very long-term dependencies} (hundreds or thousands of time steps), \textbf{LSTMs typically outperform GRUs}. Some examples:
    \begin{itemize}
        \item Long text generation.
        \item Speech sequences longer than several seconds.
        \item Language models (pre-Transformer era).
        \item Medical time series with sparse temporal relations.
        \item Irregular biomechanical signals.
    \end{itemize}
    Because the cell state in an LSTM is specifically designed for long-term memory, LSTMs \textbf{tend to retain information better}. 
    
    \textbf{GRUs can forget too aggressively} because the update gate merges the input and forget gates. This coupling makes GRUs less flexible in controlling what to remember for long periods.
    

    \item \important{GRUs Can Underperform When Selective Visibility Matters}. LSTMs have an \textbf{output gate}:
    \begin{equation*}
        h_t = o_t \odot \tanh(C_t)
    \end{equation*}
    GRUs \textbf{do not}. This means \textbf{GRUs cannot selectively hide memory and it is critical in tasks where certain features need to stay internal and not be exposed to the next layer}.
\end{itemize}

\begin{figure}[!htp]
    \centering
    \begin{tikzpicture}[
        >=LaTeX, % Use default LaTeX arrows
        % Styles 
        cell/.style={ % RNN cell
            rectangle, 
            rounded corners=5mm, 
            draw,
            very thick,
            minimum height=4cm,
            minimum width=6cm
        }, 
        cellc/.style={ % RNN cells in a chain
            cell,
            on chain,
            join
        },
        input/.style={ % Input or output node
            circle,
            minimum width=2.25em,
            draw,
            fill=gray!10,
            thick
        },
        hidden/.style={ % Hidden state
        },
        func/.style={ % Functions
            rectangle,
            draw,
            inner sep=2pt,
            minimum height=0.4cm
        },
        op/.style={ % Operators
            circle,
            draw,
            inner sep=-0.5pt,
            minimum height=0.4cm,
        },
        pointwise/.style={ % Pointwise operations and functions
            fill=gray!50,
        },
        arrow/.style={
            -latex,
            very thick
        },
        arrowc1/.style={ % Arrows with rounded corners
            arrow,
            rounded corners=.25cm
        },
        arrowc2/.style={ % Arrows with big rounded corners
            arrow,
            rounded corners=.5cm
        },
        backprop/.style={ % Backpropagation arrows
            arrow,
            dashed,
            gray
        }
    ]
    
        % GRU cell
        \node [cell] at (0,0){};

        % Internal functions and operators
        \node [func] (sigr) at (-0.75,-0.75) {$\sigma$};
        \node [func] (sigz) at (0,-0.75) {$\sigma$};
        \node [func] (tanh) at (2,-0.75) {$\tanh$};
        \node [func, pointwise] (minus) at (0,0.5) {$1-$};
        \node [op, pointwise] (mulr) at (-1.5,0) {$\times$};
        \node [op, pointwise] (mulz) at (0,1.5) {$\times$};
        \node [op, pointwise] (add) at (2,1.5) {$+$};
        \node [op, pointwise] (mulh) at (2,0) {$\times$};

        % Inputs, outputs and hidden states
        \node[hidden] (ht-1) at (-4,1.5) {$\mathbf{h}_{t-1}$};
        \node[input] (x) at (-2.5,-3) {$\mathbf{x}_{t}$};
        \node[hidden] (ht) at (4,1.5) {$\mathbf{h}_t$};
        \node[input] (y) at (2.5,3) {$\text{out } \mathbf{h}_t$};

        % Arrows from input and previous hidden states h_{t-1}
        \draw [arrowc1] (ht-1) -- (mulz) -- (add) -- (ht);
        \draw [arrowc1, -] (x) |- (-2.25, -1.4) ++(sigr);
        \draw [arrowc1, -] (x) |- (-2.25, -1.75) ++(sigr);
        \draw [arrowc1, -] (x -| sigz) ++(-1, 1.6) -| (sigz);
        \draw [arrowc1, -] (x -| sigr) ++(-1.5, 1.6) -| (sigr);
        \draw [arrowc1, -] (ht-1.east) -| (mulr);
        \draw [arrowc1, -] (ht-1) -| (-2.5,0) ++(x);
        \draw [arrowc1, -] (x) ++(0,3) |- (-1,-1.4) ++(tanh);
        \draw [arrowc2, -] (x -| tanh) ++(-4.25, 1.25) -| (tanh);

        % Arrows for operation flow within LSTM gates
        \draw [arrowc1, -] (mulr) -- ++(0,-1.3);
        \draw [arrowc1, -] (mulr) ++(0,-1.55) |- (0,-1.75) ++(tanh);
        \draw [arrowc1] (sigr) |-  node[midway, shift={(-2pt,4pt)}] {$\mathbf{r}_t$} (mulr.east);
        \draw [arrowc2] (sigz) |- (mulh);
        \draw [arrowc2] (tanh) -- node[midway, shift={(12pt,1.5pt)}] {$\tilde{\mathbf{h}}_t$} (mulh);
        \draw [arrowc2] (minus) -- (mulz);
        \draw [arrowc2] (sigz) -- node[midway, left] {$\mathbf{z}_t$} (minus);
        \draw [arrowc2] (mulh) -- (add);

        % Arrows to output and current hidden states h_t
        \draw [arrowc1] (add) coordinate[auto] -|(y);
    \end{tikzpicture}
    \captionsetup{singlelinecheck=false}
    \caption[]{
        \textbf{Gated Recurrent Unit (GRU) cell.}\cite{nntikz}
        At time step $t$, the GRU receives the current input $x_t$ and the previous hidden state $h_{t-1}$, and computes the new hidden state $h_t$, which serves both as output and as recurrent input to the next time step.

        \highspace
        The \textbf{reset gate} $r_t = \sigma(W_r [h_{t-1}, x_t] + b_r)$ controls how much of the previous hidden state contributes to the candidate activation. When $r_t$ is close to $0$, past information is ignored; when it is close to $1$, past information is fully considered.

        \highspace
        The \textbf{candidate hidden state} is computed as
        \begin{equation*}
        \tilde{h}_t = \tanh\!\big(W_h [r_t \odot h_{t-1}, x_t] + b_h\big),
        \end{equation*}
        where the reset gate filters the past before combining it with the current input.

        \highspace
        The \textbf{update gate} $z_t = \sigma(W_z [h_{t-1}, x_t] + b_z)$ determines how much of the previous hidden state is retained and how much of the candidate is incorporated. The $1-$ term is a smart way to control at the same time the contribution of the past and the new candidate: $z=1$ drops the past ($1-z=0$) and fully incorporates the new candidate ($z=1$), while $z=0$ retains the past ($1-z=1$) and ignores the new candidate ($z=0$). The final hidden state is computed as
        \begin{equation*}
        h_t = (1 - z_t)\odot h_{t-1} + z_t \odot \tilde{h}_t,
        \end{equation*}
        which can be interpreted as a \textbf{learned interpolation} between past and new information.

        \highspace
        Unlike LSTMs, GRUs \textbf{do not maintain a separate cell state} $C_t$. Instead, the hidden state $h_t$ directly combines both short-term and long-term information, resulting in a \textbf{simpler and more computationally efficient architecture}.
    }
    \label{fig:gru}
\end{figure}