\paragraph{Priority Scatter}

\definition{Priority Scatter} is a collision resolution strategy where \textbf{each element in the input array is assigned a priority value} based on its position (or another criterion). When multiple threads attempt to write to the same memory location, the \textbf{element with the highest priority wins}, and the others are discarded.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{star} \textbf{Key Features \& Advantages}}
\end{flushleft}
\begin{itemize}
    \item \textbf{Predefined Priority System}. \hl{Each element has a priority}, which could be based on position, timestamp, depth, or another factor.

    \item[\textcolor{Green3}{\faIcon{check}}] \textcolor{Green3}{\textbf{Deterministic Resolution}}. Unlike Atomic Scatter, where race conditions make the final result unpredictable, Priority Scatter \textbf{ensures a well-defined winner}.
    
    \item[\textcolor{Green3}{\faIcon{check}}] \textcolor{Green3}{\textbf{Common in 3D Graphics}}. In 3D rendering, when multiple objects overlap at the same pixel, the object closest to the camera (highest priority) is displayed.
\end{itemize}

\highspace
\begin{flushleft}
    \textcolor{Red2}{\faIcon{times-circle} \textbf{Disadvantages}}
\end{flushleft}
\begin{itemize}[label=\textcolor{Red2}{\faIcon{times}}]
    \item \important{Loss of Information}. Lower-priority elements are discarded entirely.
    \item \important{Predefined Priority Needed}. Requires an external ranking mechanism.
    \item \important{Not Always Fair}. Some data might consistently dominate, suppressing important contributions.
\end{itemize}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{balance-scale} \textbf{Comparison Atomic vs. Merge vs. Priority Scatter}}
\end{flushleft}
\begin{table}[!htp]
    \centering
    \begin{tabular}{@{} l l l @{}}
        \toprule
        Strategy & \textbf{Resolution Mechanism} & \textbf{Result Type} \\
        \midrule
        \textbf{Atomic}     & First write wins (random)     & Non-deterministic \\
        \textbf{Merge}      & Merges values (e.g. \texttt{sum}/\texttt{max})  & Aggregated \\
        \textbf{Priority}   & Highest-priority wins         & Deterministic \\
        \bottomrule
    \end{tabular}
    \caption{Differences between Scatter strategies.}
\end{table}
