\subsubsection{Probabilistic Formulation}

Suppose we are given an input sequence:
\begin{equation*}
    x = \left(x_1, x_2, \dots, x_m\right)
\end{equation*}
And we want to produce an output sequence:
\begin{equation*}
    y = \left(y_1, y_2, \dots, y_n\right)
\end{equation*}
The goal of sequence-to-sequence learning is to create a model that can predict an output sequence based on an input sequence, even when the two sequences have different lengths (i.e., $m \neq n$).

\highspace
\begin{flushleft}
    \textcolor{Red2}{\faIcon{times-circle} \textbf{\Naive Approach: predict the whole sequence at once}}
\end{flushleft}
A \naive approach would be to treat the entire output sequence as a single target and try to predict it in one shot. However, this is not practical because the output is a \textbf{structured object} (a sequence), not a single label. The number of possible sequences grows exponentially with their length, and many different sequences may be valid outputs for the same input (e.g., multiple correct translations of a sentence).

\highspace
For this reason, instead of predicting a single output directly, we aim to model how \emph{likely} a candidate output sequence is given the input.

\highspace
Formally, the model learns a \textbf{conditional probability distribution}:
\begin{equation}
    P\left(y \mid x\right)
\end{equation}
That is:
\begin{equation*}
    P\left(y_1, y_2, \dots, y_n \mid x_1, x_2, \dots, x_m\right)
\end{equation*}
Intuitively, this measures how plausible a given output sequence is for a specific input sequence. For example, if $x$ is an English sentence and $y$ is a French sentence, then $P\left(y \mid x\right)$ reflects how likely $y$ is as a translation of $x$.

\highspace
Thus, Seq2Seq learning is not a deterministic mapping from input to output, but a \textbf{probabilistic modeling problem}.

\highspace
\textcolor{Green3}{\faIcon{book} \textbf{Prediction rule.}} Once the model has learned this distribution, the predicted output is the sequence with the highest probability:
\begin{equation}
    y^{*} = \arg\max_{y} P\left(y \mid x\right)
\end{equation}
That is, among all possible output sequences, we select the most likely one given the input.

\newpage

\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{What is $P\left(y \mid x\right)$ really?}}
\end{flushleft}
Forget neural networks for a second. The expression $P\left(y \mid x\right)$ means: ``\emph{in the \textbf{real world}, how likely is output $y$ given input $x$?}''. For example, if $x$ is ``\emph{I am tired}'', the possible outputs $y$ could be ``\emph{Je suis fatigué}'' (which means ``I am tired'' in French) or ``\emph{Je mange une pomme}'' (which means ``I eat an apple'' in French). The first output is much more likely than the second one, so $P\left(y \mid x\right)$ should assign a higher probability to the first translation.

\highspace
This simple example shows that the quantity $P\left(y \mid x\right)$ represents a \textbf{true underlying distribution} between inputs and outputs in the real world. We can call this distribution $P_{\text{true}}\left(y \mid x\right)$, and it is determined by the data-generating process (e.g., natural language).

\highspace
In practice, it is impossible to know $P_{\text{true}}\left(y \mid x\right)$ exactly because we only have access to a finite dataset of examples and can only observe a limited number of input-output pairs.\footnote{%
    The reason why we observe only a limited number of input-output pairs is that the true distribution $P_{\text{true}}\left(y \mid x\right)$ is defined over a very large (often effectively infinite) space of possible inputs and outputs. In practice, it is impossible to observe all possible combinations. Instead, we only have access to a finite dataset of samples drawn from this distribution.

    \highspace
    More formally, we observe pairs:
    \begin{equation*}
        \left(x^{(1)}, y^{(1)}\right), \left(x^{(2)}, y^{(2)}\right), \dots, \left(x^{(N)}, y^{(N)}\right)
    \end{equation*}
    Which can be seen as samples from the joint distribution $P_{\text{true}}(x, y)$.

    \highspace
    For example, in natural language, the number of possible sentences is extremely large, and we can only observe a very small subset of them in our dataset.%
} So we only observe a finite set of input-output pairs:
\begin{equation*}
    \left(x^{(1)}, y^{(1)}\right), \left(x^{(2)}, y^{(2)}\right), \dots, \left(x^{(N)}, y^{(N)}\right) \sim P_{\text{true}}\left(x, y\right)
\end{equation*}
Therefore, we must learn an approximation of $P_{\text{true}}\left(y \mid x\right)$ based on the observed data.

\highspace
To this end, we introduce a parametric model:
\begin{equation*}
    P\left(y \mid x; \theta\right)
\end{equation*}
Where $\theta$ denotes the trainable parameters of the model (e.g., the weights of a neural network). This defines a \textbf{family of possible distributions}, and different values of $\theta$ correspond to different models.

\highspace
The goal of training is to find parameters $\theta$ such that:
\begin{equation*}
    P\left(y \mid x; \theta\right) \approx P_{\text{true}}\left(y \mid x\right)
\end{equation*}
That is, the model assigns high probability to correct output sequences and low probability to incorrect ones.