\subsubsection{No-Signaling Principle}\label{subsubsection: no-signaling principle}

The \definition{No-Signaling Theorem} asserts that: \textbf{no information can be transmitted faster than the speed of light, even using entangled quantum systems}.

\highspace
Although \textbf{entanglement leads to instantaneous correlations} between distant qubits, it \textbf{does not allow communication} or information transfer without a classical channel.

\highspace
This principle \textbf{preservers causality} and ensures that quantum mechanics remains compatible with \textbf{special relativity}.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{check-circle} \textbf{Explained Proof}}
\end{flushleft}
To understand the \textbf{no-signal principle} in quantum mechanics, consider the classical scenario involving two observers, Alice and Bob, who share an \textbf{entangled pair} of qubits. Specifically, they share the Bell state:
\begin{equation*}
    \ket{\Phi^{+}} = \dfrac{1}{\sqrt{2}}\left(\ket{00} + \ket{11}\right)
\end{equation*}
Alice has qubit $a$, Bob has qubit $b$, and they are located far apart, even at opposite ends of the universe.

\highspace
\textcolor{Green3}{\textbf{What happens when Alice measures?}} Suppose Alice chooses to \textbf{measure her qubit} in the computational basis.
\begin{itemize}
    \item If she gets outcome \ket{0}, the join state \textbf{collapses} to \ket{00}
    \item If she gets \ket{1}, the state collapses to \ket{11}
\end{itemize}
In both cases, Bob's qubit becomes \textbf{perfectly correlated} with Alice's result. \textbf{So yes}, Alice's measurement causes the collapse of Bob's qubit; instantly, no matter the distance.

\highspace
\textcolor{Green3}{\textbf{Can Bob detect this?}} Here's the crucial point: \textbf{Bob cannot tell whether Alice has measured her qubit}.

\highspace
Why? Because from \textbf{Bob's perspective}, before doing anything, his own qubit is described by the \textbf{reduced density matrix}:
\begin{equation*}
    \rho_{B} = \dfrac{1}{2}\ket{0}\bra{0} + \dfrac{1}{2}\ket{1}\bra{1}
\end{equation*}
This is a \textbf{completely mixed state}. If Bob measures it, he gets:
\begin{itemize}
    \item 50\% chance of \ket{0}
    \item 50\% chance of \ket{1}
\end{itemize}
Now suppose \textbf{Alice has already measured}, and the state is now \ket{00}.
\begin{itemize}
    \item Bob's qubit is now \textbf{definitely \ket{0}}
    \item But Bob doesn't know that, and he has \textbf{no way to know}
    \item The only way for him to learn this is:
    \begin{enumerate}
        \item \textbf{To measure} (and discover the value)
        \item Or \textbf{to be told} by Alice via classical communication
    \end{enumerate}
\end{itemize}

\highspace
\textcolor{Green3}{\textbf{Why this preserves No-Signaling}}. Even though \textbf{Alice's measurement affects the state} of Bob's qubit, \textbf{it does not change what Bob sees statistically}, unless he receives a message from her. Therefore:
\begin{itemize}
    \item Alice \textbf{cannot use entanglement} to send messages faster than light.
    \item Bob's observable outcomes are \textbf{indistinguishable} whether Alice measured or not.
    \item \textbf{Entanglement creates correlations}, not communication.
\end{itemize}

\highspace
\textcolor{Green3}{\textbf{Summary of the story}}
\begin{itemize}
    \item Entanglement causes \textbf{instant collapse of joint state} when one qubit is measured.
    \item But the \textbf{other party cannot detect} this collapse unless they \textbf{measure} or are \textbf{told}.
    \item No observable effect on Bob's side \texttt{=} \textbf{no information transfer} \texttt{=} no violation of causality.
\end{itemize}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{Why it matters?}}
\end{flushleft}
The no-signaling principle \textbf{guarantees}:
\begin{itemize}[label=\textcolor{Green3}{\faIcon{check}}]
    \item \textbf{No faster-than-light communication}, preserving consistency with relativity
    \item \textbf{No ``quantum telepathy''}, even in systems that exhibit entanglement
    \item Measurement outcomes are \textbf{random locally}, even when entanglement exists globally.
\end{itemize}