\subsubsection{Hadamard Gate (H)}\label{subsubsection: Hadamard Gate H}

The \definitionWithSpecificIndex{Hadamard Gate (H)}{Single Qubit Gates: Hadamard Gate (S)}{} \textbf{changes the basis representation of the qubit}. More concretely:
\begin{equation*}
    H : \left\{ \ket{0}, \, \ket{1}\right\} \longleftrightarrow \left\{ \ket{+}, \, \ket{-}\right\}
\end{equation*}
It maps computational basis states into equal superpositions and vice versa, allowing phase information to be transformed into measurable probabilities.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{square-root-alt} \textbf{Matrix Representation}}
\end{flushleft}
\begin{equation}
    H = \dfrac{1}{\sqrt{2}} 
    \begin{pmatrix}
        1 & 1 \\ 1 & -1
    \end{pmatrix}
\end{equation}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Action on Basis States}}
\end{flushleft}
\begin{align}
    H \ket{0} &= \dfrac{1}{\sqrt{2}}\left(\ket{0} + \ket{1}\right) = \dfrac{\ket{0} + \ket{1}}{\sqrt{2}} = \ket{+} \\[.5em]
    H \ket{1} &= \dfrac{1}{\sqrt{2}}\left(\ket{0} - \ket{1}\right) = \dfrac{\ket{0} - \ket{1}}{\sqrt{2}} = \ket{-}
\end{align}
The $\ket{+}$ and $\ket{-}$ \textbf{states} are known as the \definition{Hadamard basis}, or \textbf{superposition states}, with \textbf{equal probability amplitudes} for $\ket{0}$ and $\ket{1}$.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{Bloch Sphere Interpretation}}
\end{flushleft}
On the Bloch sphere, the \textbf{Hadamard gate performs two sequential rotations}:
\begin{enumerate}
    \item Rotation around the $y$-axis by $\frac{\pi}{2}$ radians ($90\degree$).
    \item Followed by a rotation around the $x$-axis by $\pi$ radians ($180\degree$).
\end{enumerate}
This brings $\ket{0}$ to the $+x$-axis ($\ket{+}$) and $\ket{1}$ to the $-x$-axis ($\ket{-}$). As a result, \textbf{qubit moves from pole to equator}, entering \textbf{maximum superposition}.

\begin{figure}[!htp]
    \centering
    \tdplotsetmaincoords{70}{120}
    \begin{tikzpicture}[tdplot_main_coords, scale=3.4, >=Latex]

        \def\r{1}

        % Axes
        \draw[->, very thick, Green3] (-1.25,0,0) -- (1.35,0,0) node[anchor=north east] {$x$};
        \draw[->, thick] (0,0,0) -- (0,1.25,0) node[anchor=west] {$y$};
        \draw[->, thick] (0,0,-1.2) -- (0,0,1.35) node[anchor=south] {$z$};
        \draw[dashed, thin, black!45] (0,-1.10,0) -- (0,0,0);

        % Bloch sphere
        \tdplotdrawarc[dashed]{(0,0,0)}{\r}{0}{180}{}{}
        \tdplotdrawarc[thick]{(0,0,0)}{\r}{180}{360}{}{}

        \tdplotsetrotatedcoords{90}{0}{90}
        \tdplotdrawarc[tdplot_rotated_coords,dashed]{(0,0,0)}{\r}{0}{180}{}{}
        \tdplotdrawarc[tdplot_rotated_coords,thick]{(0,0,0)}{\r}{180}{360}{}{}
        \tdplotsetrotatedcoords{0}{0}{0}

        % States
        \fill (0,0,1) circle (0.018);
        \node[anchor=south east] at (0,0,1.08) {$\ket{0}$};

        \fill (0,0,-1) circle (0.018);
        \node[anchor=north east] at (0,0,-1.08) {$\ket{1}$};

        \fill[Green3] (1,0,0) circle (0.018);
        \node[Green3, anchor=west] at (1.23,0.10,0) {$\ket{+}$};

        \fill[Green3] (-1,0,0) circle (0.018);
        \node[Green3, anchor=east] at (-1.18,0,0.05) {$\ket{-}$};

        % Transformation arrows
        \draw[->, very thick, magenta]
            (0,0,1)
            .. controls (0.35,0,0.75) and (0.75,0,0.35) ..
            (1,0,0);

        \draw[->, very thick, magenta]
            (0,0,-1)
            .. controls (-0.35,0,-0.75) and (-0.75,0,-0.35) ..
            (-1,0,0);

        % Labels
        \node[magenta, anchor=south east] at (0.45,0,0.55) {$H\ket{0}=\ket{+}$};
        \node[magenta, anchor=north west] at (-0.45,0,-0.55) {$H\ket{1}=\ket{-}$};
    \end{tikzpicture}
    \caption{Bloch sphere interpretation of the Hadamard gate. The gate maps $\ket{0}$ to $\ket{+}$ and $\ket{1}$ to $\ket{-}$, moving computational basis states from the $z$-axis to the $x$-axis.}
    \label{fig:hadamard-bloch}
\end{figure}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Superposition and Measurement}}
\end{flushleft}
After applying $H$ to $\ket{0}$, the resulting qubit is:
\begin{equation*}
    \ket{\psi} = \dfrac{1}{\sqrt{2}}\left(\ket{0} + \ket{1}\right)
\end{equation*}
This state is a \textbf{superposition} of $\ket{0}$ and $\ket{1}$, meaning that the qubit does not have a definite classical value prior to measurement. Instead, it is described by \textbf{probability amplitudes} associated with each basis state.

\newpage

\noindent
Upon measurement in the computational basis, the state collapses probabilistically:
\begin{itemize}
    \begin{multicols}{2}
        \item \textbf{Probability of 0}:
        \begin{equation*}
            \left|\dfrac{1}{\sqrt{2}}\right|^{2} = \dfrac{1}{2}
        \end{equation*}
        \item \textbf{Probability of 1}:
        \begin{equation*}
            \dfrac{1}{2}
        \end{equation*}
    \end{multicols}
\end{itemize}
After the measurement:
\begin{itemize}
    \item The qubit collapses to $\ket{0}$ with probability $\frac{1}{2}$, or to $\ket{1}$ with probability $\frac{1}{2}$.
    \item The original superposition is \textbf{destroyed}, and cannot be recovered.
\end{itemize}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Eigenvalues and Eigenstates}}
\end{flushleft}
The Hadamard operator has eigenvalues:
\begin{equation*}
    \lambda = +1, \hspace{2em} \lambda = -1
\end{equation*}
with corresponding eigenstates given by the solutions of:
\begin{equation*}
    H\ket{v} = \lambda \ket{v}
\end{equation*}
These eigenstates are superpositions of $\ket{0}$ and $\ket{1}$, and are aligned along an axis halfway between the $x$ and $z$ directions on the Bloch sphere. Precisely, the eigenvectors are:
\begin{align*}
    \ket{v_{+}} &= \cos\left(\dfrac{\pi}{8}\right)\ket{0} + \sin\left(\dfrac{\pi}{8}\right)\ket{1} \\
    \ket{v_{-}} &= \sin\left(\dfrac{\pi}{8}\right)\ket{0} - \cos\left(\dfrac{\pi}{8}\right)\ket{1}
\end{align*}
Where $\ket{v_{+}}$ corresponds to the eigenvalue $+1$ and $\ket{v_{-}}$ corresponds to the eigenvalue $-1$; the cosine and sine factors ensure normalization of the eigenstates.

\highspace
Unlike Pauli gates, the Hadamard eigenstates lie along a \textbf{diagonal axis between the computational basis $z$ and the Hadamard basis $x$}, reflecting its role as a \textbf{basis transformation} rather than a simple bit or phase flip.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{list-ul} \textbf{Key Properties}}
\end{flushleft}
\begin{enumerate}
    \item \textbf{Unitary}: $H^\dagger H = I$, so normalization is preserved.
    
    \item \textbf{Hermitian}: $H^\dagger = H$, therefore it corresponds to an observable.
    
    \item \textbf{Self-inverse}: $H^2 = I$. Applying the Hadamard gate twice returns the original state.
    
    \item \textbf{Basis transformation}: H maps the computational basis $\left\{\ket{0}, \ket{1}\right\}$ to the Hadamard basis $\left\{\ket{+}, \ket{-}\right\}$:
    \begin{equation*}
        H\ket{0} = \ket{+}, \hspace{2em} H\ket{1} = \ket{-}
    \end{equation*}
    
    \item \textbf{Interference enabler}: $H$ converts phase differences into amplitude differences, making them observable through measurement.
    
    \item \textbf{Axis mixing}: $H$ transforms Pauli operators as:
    \begin{equation*}
        HZH = X, \hspace{2em} HXH = Z
    \end{equation*}
\end{enumerate}