\section{Multigrid methods}

\subsection{Idea of MG methods}

The \definition{Multigrid (MG) methods} are efficient \textbf{algorithms for solving large systems of linear equations}, particularly those \textbf{arising from the discretization of partial differential equations} (PDEs). They're especially useful for problems that exhibit behavior on multiple scales.

\highspace
A multigrid (MG) method is an iterative algorithm of the form:
\begin{equation}
    \mathbf{x}^{\left(k+1\right)} = \mathrm{MG}\left(\mathbf{x}^{\left(k\right)}\right) \hspace{2em} k \ge 0
\end{equation}
For solving the (typically) sparse linear systems of equations stemming from the numerical discretization of differential equations. The \textbf{MG methods are based on}:
\begin{itemize}
    \item \textbf{\emph{Hierarchy of levels}} (associated with a hierarchy of discretization):
    \begin{itemize}
        \item \textbf{\underline{Fine Grid}}. The finest grid \textbf{captures the most detailed features} of the problem. This is where the original problem is defined and where the final solution needs to be accurate.
        
        \item \textbf{\underline{Coarse Grids}}. They are lower resolution versions of the fine grid. They \textbf{capture broader}, \textbf{large-scale features} of the problem. The coarser the grid, the \textbf{fewer the details}, but \textbf{computations are cheaper and faster}.
        
        Coarse grids help in correcting the errors that are hard to eliminate on finer grids due to their global nature.
    \end{itemize}

    \item \textbf{\emph{MG cycles}} reduce all error components by a fixed amount (bounded well below one), regardless of the dimension $n$ of the system.
\end{itemize}
The \textbf{main idea} of MG is to \textbf{accelerate the convergence of a basic iterative method} by a global correction of the fine grid solution approximation accomplished by solving a coarse problem. The coarse-level problem should be \emph{similar} to the fine grid problem. The cost of (direct) solution of the coarse problem should be negligible compared to the cost of one relaxation sweep on the fine grid.

\highspace
In other words, the main goal of the multigrid method is to speed up the convergence of an iterative method for solving systems of linear equations. This acceleration is achieved by globally correcting the solution approximation on the fine grid by solving a similar problem on a coarser grid.