\section{Multiprocessors}

\subsection{What is a multiprocessor?}

A \definition{Multiprocessor} is a computer system with \textbf{tightly coupled processors}, coordinated by a \textbf{single operating system}, and typically offering a \textbf{shared memory abstraction} (shared address space). The idea is that \hl{processors work in parallel} but \hl{remain part of one coherent machine}, not a loose cluster. The execution paradigm is \textbf{MIMD (Multiple Instruction, Multiple Data)}: each processor/core runs its own instruction stream on its own data (page \pageref{def: MIMD}).

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{bullseye} \textbf{Goals of Multiprocessors}}
\end{flushleft}
Multiprocessors emerged to address the limits of uniprocessor performance:
\begin{enumerate}
    \item \textbf{Performance}: Achieve \emph{very high throughput} by exploiting thread-level parallelism.
    \item \textbf{Scalability}: System performance should grow (ideally linearly) as more processors are added.
    \item \textbf{Reliability}: Faults can sometimes be tolerated thanks to redundancy.
    \item \textbf{Versatility}: Found in domains ranging from \textbf{embedded systems} to \textbf{high-end servers}.
\end{enumerate}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{stream} \textbf{Taxonomy}}
\end{flushleft}
We distinguish multiprocessors based on integration and scale:
\begin{itemize}
    \item \definition{Multicores (on-chip multiprocessors)}: Several cores on the same chip. Called \textbf{manycores} when the number of cores grows beyond 32. For example, Intel i7, AMD Ryzen, ARM bit.LITTLE.
    \item \definition{SMP - Symmetric Multiprocessors (centralized shared memory)}: Typically $\le 32$ processors. All processors share \textbf{a single centralized memory} with \textbf{Uniform Memory Access (UMA)} time. Most existing multicore chips today are SMPs. Cache coherence is handled with \textbf{snooping protocols}.
    \item \definition{Large-Scale Multiprocessors (Distributed Shared Memory)}: Needed when scaling beyond 32 processors. Memory is \textbf{physically distributed}, leading to \textbf{NUMA (Non-Uniform Memory Access)} latencies. Typically managed with \textbf{directory-based cache coherence}.
\end{itemize}

\highspace
Multiprocessors are not just ``more cores''. They are about how we \textbf{organize and coordinate processors, memory, and communication}. The shift from single-core $\rightarrow$ multicore $\rightarrow$ manycore reflects the industry response to ILP and frequency walls. The choice between \textbf{SMP vs. NUMA} reflects scalability limits: centralized memory works up to a point, beyond which distribution is necessary.