\section*{Preface}

Every theory section in these notes has been taken from two sources:
\begin{itemize}
    \item Ian Goodfellow and Yoshua Bengio and Aaron Courville, \href{https://www.deeplearningbook.org/}{Deep Learning}, MIT Press. \cite{Goodfellow-et-al-2016}
    \item Course slides.\cite{course-slides-polimi}
\end{itemize}
About:
\begin{itemize}
    \item[\faIcon{github}] \href{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/HPC-E-PoliMI-university-notes}{GitHub repository}
    \begin{center}
        \qrcode{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/HPC-E-PoliMI-university-notes}
    \end{center}
\end{itemize}
These notes are an unofficial resource and shouldn't replace the course material or any other book on artificial neural networks and deep learning. It is not made for commercial purposes. I've made the following notes to help me improve my knowledge and maybe it can be helpful for everyone.

As I have highlighted, a student should choose the teacher's material or a book on the topic. These notes can only be a helpful material.

\highspace
During the Artificial Neural Networks and Deep Learning course, me and my other three colleagues Alberto Ondei, \href{https://it.linkedin.com/in/javedabdullah}{Abdullah Javed} and \href{https://www.hpc.cineca.it/staff/barbari-filippo/}{Filippo Barbari}, we created two projects:
\begin{itemize}
    \item[\faIcon{github}] \href{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/AN2DL-Challenge-1}{Time Series Classification}. Deep Neural Netowrk (TCN $+$ BiLSTM with Attention) model on multivariate time-series classification.
    \begin{center}
        \qrcode{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/AN2DL-Challenge-1}
    \end{center}
    \item[\faIcon{github}] \href{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/AN2DL-Challenge-2}{Image Classification}. Histopathology image classification to predict molecular subtypes.
    \begin{center}
        \qrcode{https://github.com/PoliMI-HPC-E-notes-projects-AndreVale69/AN2DL-Challenge-2}
    \end{center}
\end{itemize}