\section{Single Qubit Gates}

\subsection{Operations on Qubits}

Quantum computing relies on three essential operations:
\begin{enumerate}
    \item \textbf{Logic Gates}: transform quantum states.
    \item \textbf{Measurements}: extract classical information from qubits.
    \item \textbf{Initialization Procedures}: prepare qubits in known starting states.
\end{enumerate}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Logic Gates: Reversible Transformations}}
\end{flushleft}
Quantum \definitionWithSpecificIndex{Logic Gates}{Single Qubit Gates: Logic Gates}{} are the building blocks of quantum circuits. They \textbf{manipulate the state of qubits in a controlled and deterministic way}, analogous to classical logic gates (like AND or NOT), but with key quantum differences:
\begin{itemize}
    \item \important{Operate on Single or Multiple Qubits}. They can affect one qubit (single-qubit gates) or multiple qubits simultaneously (multi-qubit gates):
    \begin{itemize}
        \item \textbf{Single-Qubit Gates}: These gates, such as the Pauli-X, Y, Z,\break Hadamard (H), and Phase (S) gates, operate on a single qubit to change its state. We will explore these gates in detail in the next sections.
        \item \textbf{Multi-Qubit Gates}: These gates, such as the CNOT (Controlled-NOT) and Toffoli gates, operate on multiple qubits and can create \emph{entanglement} (we will discuss this in detail later) between them, which is a key resource in quantum computing.
    \end{itemize}


    \item \important{Reversible Transformations}. Unlike many classical gates (e.g., AND), \textbf{all quantum gates are reversible}. Mathematically, this means they are represented by \textbf{unitary matrices} $U$, satisfying:
    \begin{equation*}
        U^\dagger U = U U^\dagger = I
    \end{equation*}
    Where $U^\dagger$ is the conjugate transpose of $U$, and $I$ is the identity matrix. Because of this reversibility, quantum gates \hl{do not lose information} and the \hl{original state can be recovered} by applying $U^\dagger = U^{-1}$ to the output state. This is a fundamental difference from classical logic gates, which can be irreversible (e.g., AND gate loses information about the inputs).

    Finally, this property is related to fundamental principles such as the \textbf{no-cloning theorem} (\autopageref{thm:no-cloning}) and the \textbf{no-deleting theorem} (we cannot delete quantum information arbitrarily, since it is preserved under unitary transformations).


    \item \important{State Evolution}. If a qubit is in the state:
    \begin{equation*}
        \ket{\psi} = a\ket{0} + b\ket{1}
    \end{equation*}
    Then after applying a gate $U$, the new state becomes:
    \begin{equation*}
        \ket{\psi'} = U\ket{\psi} = aU\ket{0} + bU\ket{1}
    \end{equation*}
    This \textbf{transformation is deterministic} and \textbf{preserves the normalization of the state} (i.e., $\left|a\right|^2 + \left|b\right|^2 = 1$ remains true after applying $U$). The specific effect of the gate depends on its matrix representation, which we will explore in the next sections.
\end{itemize}
We can think of quantum logic gates as \textbf{rotations on the Bloch sphere}, smoothly transforming the qubit's state without destroying its quantum nature.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Measurement: Irreversible Collapse of State}}
\end{flushleft}
\definitionWithSpecificIndex{Measurement}{Single Qubit Gates: Measurement}{} is the process of extracting information from a qubit. Unlike logic gates, measurement is \textbf{irreversible} and fundamentally probabilistic.
\begin{itemize}
    \item \important{Irreversible Process}. Once a qubit is measured, its original quantum state is \textbf{lost}. This distinguishes measurement from unitary gate operations.
    
    
    \item \important{State Collapse}. Measuring a qubit in the computational basis $\left\{\ket{0}, \ket{1}\right\}$ causes the state:
    \begin{equation*}
        \ket{\psi} = a\ket{0} + b\ket{1}
    \end{equation*}
    To \textbf{collapse} to:
    \begin{itemize}
        \item $\ket{0}$ with probability $\left|a\right|^2$.
        \item $\ket{1}$ with probability $\left|b\right|^2$.
    \end{itemize}
    
    
    \item \important{Extraction of Classical Information}. The measurement outcome is a \textbf{classical bit} (0 or 1). After measurement, subsequent measurements in the same basis yield the same result with probability 1, since the state has collapsed to a definite state.
\end{itemize}
Measurement can be described using \textbf{projection operators}:
\begin{equation*}
    M_0 = \ket{0}\bra{0} \quad M_1 = \ket{1}\bra{1}
\end{equation*}
The probability of obtaining outcome $k$ (where $k \in \{0,1\}$) when measuring $\ket{\psi}$ is given by:
\begin{align*}
    p_k =& \bra{\psi} M_k \ket{\psi} \\[.2em]
    \downarrow& \text{ substitute } M_k = \ket{k}\bra{k} \\[.2em]
    =& \bra{\psi} \left(\ket{k}\bra{k}\right) \ket{\psi} \\[.2em]
    \downarrow& \text{ using the associativity of inner products} \\[.2em]
    =& \left(\braket{\psi}{k}\right) \left(\braket{k}{\psi}\right) \\[.2em]
    \downarrow& \text{ since } \braket{\psi}{k} = \left(\braket{k}{\psi}\right)^* \\[.2em]
    & \text{ where * denotes complex conjugation} \\[.2em]
    =& \left(\braket{k}{\psi}\right)^{*} \left(\braket{k}{\psi}\right) \\[.2em]
    \downarrow& \text{ which is the definition of squared magnitude, \autopageref{sec:magnitude}} \\[.2em]
    =& \left|\braket{k}{\psi}\right|^2
\end{align*}
In other words, the probability of measuring outcome $k$ is the \textbf{squared magnitude of the amplitude} of $\ket{\psi}$ along the direction $\ket{k}$. And the post-measurement state of the qubit, conditioned on obtaining outcome $k$, becomes:
\begin{equation*}
    \ket{\psi_k} = \frac{M_k \ket{\psi}}{\sqrt{p_k}}
\end{equation*}
Measurement is like \textbf{asking a question} to the qubit: ``are you in state $\ket{0}$ or $\ket{1}$?''. The answer is classical (i.e., 0 or 1), but the act of asking \textbf{changes the system}.

\highspace
\begin{examplebox}[: Calculating Measurement Probabilities]
    Consider a qubit in the state:
    \begin{equation*}
        \ket{\psi} = a \ket{0} + b \ket{1}
    \end{equation*}
    With $\left|a\right|^2 + \left|b\right|^2 = 1$.
    \begin{itemize}
        \item Probability of measuring $0$. Using the projector:
        \begin{equation*}
            M_0 = \ket{0}\bra{0}
        \end{equation*}
        We compute:
        \begin{align*}
            p_0 =& \bra{\psi} M_0 \ket{\psi} \\[.2em]
            =& \braket{\psi}{0} \braket{0}{\psi} \\[.2em]
            =& \left(\braket{\psi}{0}\right) \left(\braket{0}{\psi}\right) \\[.2em]
            \downarrow& \text{ since } \braket{\psi}{0} = \left(\braket{0}{\psi}\right)^* \\[.2em]
            =& \left|\braket{0}{\psi}\right|^2 \\[.2em]
            =& \left|a\right|^2
        \end{align*}


        \item Probability of measuring $1$. Using the projector:
        \begin{equation*}
            M_1 = \ket{1}\bra{1}
        \end{equation*}
        We compute:
        \begin{align*}
            p_1 =& \bra{\psi} M_1 \ket{\psi} \\[.2em]
            =& \braket{\psi}{1} \braket{1}{\psi} \\[.2em]
            =& \left(\braket{\psi}{1}\right) \left(\braket{1}{\psi}\right) \\[.2em]
            \downarrow& \text{ since } \braket{\psi}{1} = \left(\braket{1}{\psi}\right)^* \\[.2em]
            =& \left|\braket{1}{\psi}\right|^2 \\[.2em]
            =& \left|b\right|^2
        \end{align*}
    \end{itemize}
\end{examplebox}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Initialization: Preparing Known States}}
\end{flushleft}
The \definitionWithSpecificIndex{Initialization}{Single Qubit Gates: Initialization}{} is the process of preparing a qubit in a \textbf{known and controlled state}, typically $\ket{0}$ or $\ket{1}$. This step is essential because quantum algorithms require a well-defined starting point. It is implemented through various methods, including:
\begin{itemize}
    \item \important{Physical Preparation}. Many quantum technologies naturally relax to the ground state $\ket{0}$. Additional gates (e.g., Pauli-X) can be applied to obtain $\ket{1}$ if needed.
    \item \important{Initialization via Measurement}. Measuring a qubit projects it into a basis state. If the outcome is not the desired one, a corrective gate can be applied. For example, if measurement yields $\ket{1}$ but $\ket{0}$ is needed, apply an X gate to flip it to $\ket{0}$.
\end{itemize}
Initialization is the \textbf{first step} in any quantum computation: (1) \textbf{prepare} qubits in known states, (2) \textbf{apply} them using quantum gates, and (3) \textbf{measure} to obtain classical results. It is analogous to \textbf{resetting a classical register} before starting a computation in classical computing.