\subsection{Quantum Logic Gates Overview}

In quantum computing, quantum logic gates are the primary tools used to \textbf{manipulate qubits}. These gates perform \textbf{unitary transformations} on the state of one or more qubits, meaning they \textbf{preserve the norm} of the quantum state and are \textbf{reversible}. The \hl{design} of \hl{quantum algorithms} and the \hl{execution of quantum circuits rely entirely on the sequential application of these gates}.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{list-ul} \textbf{Types of Quantum Gates}}
\end{flushleft}
\definition{Quantum Gates} can be classified into:
\begin{itemize}
    \item \definition{Single-Qubit Gates}: These gates operate on \textbf{individual qubits}, modifying their state in isolation.
    \item \textbf{Multiple-Qubit Gates}: These gates act on \textbf{two or more qubits simultaneously}, allowing for entanglement and complex correlations between qubits.
\end{itemize}
In this section we will focus on single qubit gates.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{square-root-alt} \textbf{Mathematical Framework: Qubit as Vectors, Gates as Matrices}}
\end{flushleft}
To describe quantum superposition mathematically, we require a framework that supports linear combinations of states, probabilities derived from inner products, complex amplitudes and phases, and linear state transformations. The mathematical structure satisfying all these requirements is a \textbf{complex Hilbert space}. Thus, a qubit is naturally represented as a \textbf{unit vector} in a two-dimensional complex Hilbert space $\mathbb{C}^{2}$.

\highspace
Using the Dirac notation, the basis states $\ket{0}$ and $\ket{1}$ form an orthonormal basis for this space, satisfying $\braket{0}{0} = \braket{1}{1} = 1$ and $\braket{0}{1} = \braket{1}{0} = 0$. Any qubit state can be expressed as a linear combination of these basis states:
\begin{equation*}
    \ket{\psi} = a\ket{0} + b\ket{1} \hspace{2em} \text{with } \left|a\right|^{2} + \left|b\right|^{2} = 1
\end{equation*}
Two state vectors that differ only by a global phase factor $e^{i\phi}$ represent the same physical qubit, since measurement probabilities remain unchanged (see \autopageref{sec:global-phase-equivalence}).

\highspace
Choosing the computational basis $\{\ket{0}, \ket{1}\}$, we can represent these states as column vectors:
\begin{equation*}
    \ket{0} = \begin{bmatrix}
        1 \\ 0
    \end{bmatrix}
    \qquad
    \ket{1} = \begin{bmatrix}
        0 \\ 1
    \end{bmatrix}
\end{equation*}
Therefore, the qubit state can be written as:
\begin{equation*}
    \ket{\psi} = a\ket{0} + b\ket{1} = \begin{bmatrix}
        a \\ b
    \end{bmatrix}
\end{equation*}
This vector representation allows us to apply linear algebra techniques to analyze and manipulate qubits.

\highspace
Once qubits are represented as vectors, transformations of these states must be \textbf{linear operators} acting on the vector space. In a chosen basis, linear operators are represented by \textbf{matrices}. To preserve normalization and probabilities, these matrices must be \textbf{unitary}:
\begin{equation*}
    U^{\dagger} U = U U^{\dagger} = I
\end{equation*}
Thus, quantum gates -- the \textbf{building blocks of quantum circuits} -- are represented as unitary $2 \times 2$ matrices that linearly transform the state vector:
\begin{equation*}
    \ket{\psi'} = U\ket{\psi}
\end{equation*}
\textcolor{Green3}{\faIcon{question-circle} \textbf{Why Matrices?}} Because quantum mechanics is inherently linear, any physical evolution of a closed quantum system must be described by a \textbf{linear operator}. When expressed in a chosen basis (such as the computational basis $\{\ket{0}, \ket{1}\}$), this operator takes the form of a matrix.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon[regular]{star} \textbf{Unitary: The Fundamental Constraint}}
\end{flushleft}
A matrix $U$ is \textbf{unitary} if it satisfies the condition:
\begin{equation}
    U^{\dagger} U = U U^{\dagger} = I
\end{equation}
Where:
\begin{itemize}
    \item $U^{\dagger}$ is the \textbf{Hermitian transpose} (conjugate transpose) of $U$.
    \item $I$ is the \textbf{identity matrix}.
\end{itemize}
This property ensures that $U^{-1} = U^{\dagger}$, meaning every quantum operation is \textbf{reversible}, and the transformation preserves the \textbf{norm} of the state vector. So, quantum gate \textbf{\underline{must be unitary}} to ensure that the evolution of quantum states is consistent with the principles of quantum mechanics (i.e., linearity, reversibility, and preservation of probabilities).

\highspace
A \hl{general single-qubit unitary matrix} can be expressed as:
\begin{equation}
    U = \begin{pmatrix}
        \alpha & \beta \\
        \gamma & \delta
    \end{pmatrix}
    \quad \text{with} \quad \alpha, \beta, \gamma, \delta \in \mathbb{C} \quad \text{and} \quad U^{\dagger} U = I
\end{equation}

\highspace
\begin{flushleft}
    \textcolor{Red2}{\faIcon{question-circle} \textbf{Why must gates be unitary?}}
\end{flushleft}
\begin{enumerate}
    \item \definition{No-Cloning Theorem}: It is impossible to create an identical copy of an unknown qubit. Unitary transformations respect this by not duplicating information.
    \item \definition{No-Deleting Theorem}: We cannot delete quantum information arbitrarily. Since unitary gates are invertible, they do not erase information.
    \item \textbf{Preservation of Probability}: The normalization condition $\left|a\right|^{2} + \left|b\right|^{2} = 1$ must hold after any operation, and \hl{only unitary matrices preserve this norm.}
\end{enumerate}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Physical Intuition}}
\end{flushleft}
In classical circuits, logic gates like AND, OR, NOT process bits deterministically. In quantum circuits:
\begin{itemize}
    \item Gates \textbf{rotate} the quantum state on the \textbf{bloch sphere}.
    \item These rotations correspond to \textbf{unitary transformations}.
    \item The \textbf{physical implementation} of quantum gates may vary across hardware (e.g., superconducting qubits, trapped ions), but \textbf{mathematically}, the \textbf{same unitary matrix describes the gate}.
\end{itemize}
\textbf{Quantum gates do not correspond to physical gates} in the classical sense. They are \textbf{abstract operations} realized by \textbf{controlled interactions} in quantum systems.
