\subsection{Single-Qubit Quantum Circuits}

A \definition{Quantum Circuit} is a \textbf{mathematical model} for quantum computation, \textbf{composed of sequential operations applied to qubits}. In other words, a quantum circuit is a \textbf{model of computation} that describes \textbf{how a quantum state evolves step by step}. The operations in a quantum circuit include initializations, unitary gates, and measurements. In this section, we focus on \hl{circuits involving a single qubit}, which form the foundation for understanding multi-qubit systems.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{list-ul} \textbf{Circuit Elements}}
\end{flushleft}
A quantum circuit typically consists of three main elements:
\begin{enumerate}
    \item \important{Initialization} (initial state)
    \begin{itemize}
        \item Qubit is prepared in a \textbf{known state}, typically $\ket{0}$ or $\ket{1}$.
        \item Essential \textbf{starting point} for any computation.
    \end{itemize}

    \item \important{Gates}
    \begin{itemize}
        \item Single-qubit \textbf{unitary operations} (e.g., X, Z, H, S).
        \item Transform the qubit's state \textbf{reversibly}.
    \end{itemize}

    \item \important{Measurement}
    \begin{itemize}
        \item \textbf{Irreversible process}.
        \item Collapses the qubit to $\ket{0}$ or $\ket{1}$ with probabilities dictated by the quantum state's amplitudes.
    \end{itemize}
\end{enumerate}

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{tools} \textbf{Quantum Circuit Diagram: Visual Language}}
\end{flushleft}
\begin{itemize}
    \item \textbf{Horizontal lines} represent the \textbf{timeline of a single qubit}.
    \begin{center}
        \begin{quantikz}
            \lstick{$\ket{v_{\text{in}}}$}
            & & & & & & % #6
            \rstick{$\ket{v_{\text{out}}}$}
        \end{quantikz}
    \end{center}
    \item \textbf{Boxes} on the line represent \textbf{gates} (letters) or \textbf{measurements} (tachometer icon).
    \begin{center}
        \begin{quantikz}
            \lstick{$\ket{v_{\text{in}}}$}
            & \gate{A} & \gate{B} & \gate{\cdots} & \gate{N} & \meter{}
        \end{quantikz}
    \end{center}
    \item \textbf{Double lines} (if present) represent \textbf{classical information} (e.g., measurement result).
    \begin{center}
        \begin{quantikz}
            \lstick{$\ket{v_{\text{in}}}$}
            & \gate{A} & \gate{B} & \meter{} & \setwiretype{c}
            \rstick{$\ket{v_{\text{out}}}$}
        \end{quantikz}
    \end{center}
    \item[\textcolor{Red2}{\faIcon{exclamation-triangle}}] \hl{Left to right}: The \textbf{circuit evolves over time from left (input) to right (output)}.
\end{itemize}

\newpage

\begin{flushleft}
    \textcolor{Green3}{\faIcon{book} \textbf{Gate Sequence: Serial Gates}}
\end{flushleft}
Consider two gates A and B applied in sequence:
\begin{center}
    \begin{quantikz}
        \lstick{$\ket{v_{\text{in}}}$}
        & \gate{A} & \gate{B} &
        \rstick{$\ket{v_{\text{out}}}$}
    \end{quantikz}
\end{center}
\begin{enumerate}
    \item First apply A, then B.
    \item \textcolor{Red2}{\faIcon{exclamation-triangle} \textbf{Attention}}: the \textbf{order of matrix multiplication} is \textbf{reversed} when we write the combined effect of the gates as a single matrix. The gate that is applied \textbf{last} (B) appears \textbf{first} in the product.
    \begin{equation*}
        v_{\text{out}} = B A v_{\text{in}}
    \end{equation*}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{Why?}} Because first apply A to the input state $v_{\text{in}}$, then apply B to the result of $A v_{\text{in}}$.
    \item The \textbf{combined effect} of the two gates is equivalent to a single gate C = BA:
    \begin{equation*}
        \begin{quantikz}
            \lstick{$\ket{v_{\text{in}}}$}
            & \gate{BA} &
            \rstick{$\ket{v_{\text{out}}}$}
        \end{quantikz}
        \quad
        % \text{is equivalent to}
        \equiv
        \quad
        \begin{quantikz}
            \lstick{$\ket{v_{\text{in}}}$}
            & \gate{A} & \gate{B} &
            \rstick{$\ket{v_{\text{out}}}$}
        \end{quantikz}
    \end{equation*}
    \begin{equation*}
        v_{\text{out}} = C v_{\text{in}} = BA v_{\text{in}}
    \end{equation*}
    Obviously, if we merge the gates into a single gate $C$, we must respect the order of operations, which is why $C = BA$ and not $AB$.
\end{enumerate}

\highspace
\begin{examplebox}[: Hadamard followed by Hadamard]
    Let's compute H followed by H to $\ket{0}$:
    
    \begin{center}
        \begin{quantikz}
            \lstick{$\ket{0}$}
            & \gate{H} & \gate{H} &
            \rstick{$\ket{v_{\text{out}}}$}
        \end{quantikz}
    \end{center}

    \begin{enumerate}
        \item Apply H:
        \begin{align*}
            H\ket{0} &=
            \dfrac{1}{\sqrt{2}}
            \cdot
            \begin{pmatrix}
                1 & 1  \\
                1 & -1
            \end{pmatrix}
            \cdot
            \begin{pmatrix}
                1 \\
                0
            \end{pmatrix}
            \\[.3em]
            &=
            \dfrac{1}{\sqrt{2}}
            \cdot
            \begin{pmatrix}
                1 \\ 1
            \end{pmatrix}
            \\[.3em]
            &=
            \dfrac{1}{\sqrt{2}}
            \cdot
            \left(
            \begin{pmatrix}
                1 \\ 0
            \end{pmatrix}
            \cdot
            \begin{pmatrix}
                0 \\ 1
            \end{pmatrix}
            \right)
            \\[.3em]
            &=
            \dfrac{1}{\sqrt{2}}\left(\ket{0} + \ket{1}\right) = \ket{+}
        \end{align*}
        \item Applying H again, we \textbf{restore the original state due to the involution properties} ($H^{2} = I$):
        \begin{equation*}
            H\left(H\ket{0}\right) =
            \dfrac{1}{\sqrt{2}}
            \begin{pmatrix}
                1 & 1 \\
                1 & -1
            \end{pmatrix}
            \cdot
            \dfrac{1}{\sqrt{2}}
            \begin{pmatrix}
                1 \\
                1
            \end{pmatrix}
            =
            \dfrac{1}{2}
            \begin{pmatrix}
                2 \\
                0
            \end{pmatrix}
            =
            \begin{pmatrix}
                1 \\
                0
            \end{pmatrix}
            =
            \ket{0}
        \end{equation*}
    \end{enumerate}
\end{examplebox}
