\subsection{When Does a Gate Create Superposition?}

In this section, we will answer the question: \emph{when does a quantum gate create superposition?} More precisely, if we start from a \textbf{basis state} $\ket{0}$ or $\ket{1}$, when does a gate produce:
\begin{equation*}
    a\ket{0} + b\ket{1} \quad \text{with} \quad a,b \ne 0 \,\text{?}
\end{equation*}
\textcolor{Green3}{\faIcon{question-circle} \textbf{What does ``\emph{create superposition}'' really mean?}} It means that the output must contain \textbf{both basis states}. So:
\begin{itemize}
    \item $\begin{pmatrix}
        1 \\ 0
    \end{pmatrix}$ is not a superposition, it's just $\ket{0}$.
    \item $\begin{pmatrix}
        0 \\ 1
    \end{pmatrix}$ is not a superposition, it's just $\ket{1}$.
    \item $\begin{pmatrix}
        a \\ b
    \end{pmatrix}$ with $a,b \ne 0$ is a superposition, because it contains both $\ket{0}$ and $\ket{1}$ components. Indeed, we can write it as:
    \begin{equation*}
        a\ket{0} + b\ket{1} =
        a\begin{pmatrix}
            1 \\ 0
        \end{pmatrix} +
        b\begin{pmatrix}
            0 \\ 1
        \end{pmatrix} =
        \begin{pmatrix}
            a \\ b
        \end{pmatrix}
    \end{equation*}
\end{itemize}

\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{How does a matrix act on a basis state?}}
\end{flushleft}
The main idea is that \hl{each column of the matrix tells us where a basis vector goes.} For example, for a matrix:
\begin{equation*}
    A =
    \begin{pmatrix}
        a_{11} & a_{12} \\
        a_{21} & a_{22}
    \end{pmatrix}
\end{equation*}
The first column $\begin{pmatrix}
    a_{11} \\ a_{21}
\end{pmatrix}$ tells us where $\ket{0}$ goes, and the second column $\begin{pmatrix}
    a_{12} \\ a_{22}
\end{pmatrix}$ tells us where $\ket{1}$ goes. In general, a \hl{gate creates superposition if one of its columns has \textbf{two non-zero entries}.} 

\highspace
\begin{flushleft}
    \textcolor{Red2}{\faIcon{exclamation-triangle} \textbf{Constraint: Unitarity}}
\end{flushleft}
Not every matrix can be a valid quantum gate. Quantum gates must be \textbf{unitary}, meaning:
\begin{equation*}
    A^\dagger A = I
\end{equation*}
This imposes strong constraints on the matrix entries. In particular, the columns must be:
\begin{itemize}
    \item \textbf{Normalized}, total probability is 1: $\left\| \psi \right\|^{2} = \left|a\right|^{2} + \left|b\right|^{2} = 1$
    \item \textbf{Orthogonal}, inner product is zero: $\braket{v}{w} = 0$
\end{itemize}
As a consequence, some matrices that appear to create superposition are \textbf{not physically allowed}, because they are not unitary.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon[regular]{lightbulb} \textbf{Key Insight}}
\end{flushleft}
Superposition arises from the \textbf{mixing of basis states}. Therefore:
\begin{itemize}
    \item A gate \textbf{creates superposition} if at least one of its columns contains \textbf{two non-zero entries}.
    \item A gate \textbf{cannot create superposition} if its columns contain only one non-zero entry (as in diagonal or triangular matrices).
\end{itemize}
In particular:
\begin{itemize}
    \item Gates like $H$ (Hadamard) create superposition because they mix $\ket{0}$ and $\ket{1}$.
    \item Gates like $X$, $Y$, $Z$ or $S$ do not create superposition because they only modify phases and do not mix basis states.
\end{itemize}

\highspace
\begin{proof}[\textbf{Proof that $A$ cannot create superposition}]
    Consider the gate:
    \begin{equation*}
        A =
        \begin{pmatrix}
            a_{11} & a_{12} \\
            0      & a_{22}
        \end{pmatrix}
    \end{equation*}
    To create superposition from a basis state, at least one column of $A$ must contain two non-zero entries.

    \highspace
    Applying $A$ to $\ket{0}$ gives:
    \begin{equation*}
        A\ket{0}
        =
        \begin{pmatrix}
            a_{11} & a_{12} \\
            0      & a_{22}
        \end{pmatrix} \cdot \begin{pmatrix}
            1 \\ 0
        \end{pmatrix}
        =
        \begin{pmatrix}
            a_{11} \cdot 1 + a_{12} \cdot 0 \\
            0 \cdot 1 + a_{22} \cdot 0
        \end{pmatrix}
        =
        \begin{pmatrix}
            a_{11} \\
            0
        \end{pmatrix}
        =
        a_{11}\ket{0}
    \end{equation*}
    Therefore, $\ket{0}$ is not mapped to a superposition because the second entry is zero.

    \highspace
    Applying $A$ to $\ket{1}$ gives:
    \begin{equation*}
        A\ket{1}
        =
        \begin{pmatrix}
            a_{11} & a_{12} \\
            0      & a_{22}
        \end{pmatrix} \cdot \begin{pmatrix}
            0 \\ 1
        \end{pmatrix}
        =
        \begin{pmatrix}
            a_{11} \cdot 0 + a_{12} \cdot 1 \\
            0 \cdot 0 + a_{22} \cdot 1
        \end{pmatrix}
        =
        \begin{pmatrix}
            a_{12} \\
            a_{22}
        \end{pmatrix}
        =
        a_{12}\ket{0} + a_{22}\ket{1}
    \end{equation*}
    This would be a superposition only if both $a_{12} \neq 0$ and $a_{22} \neq 0$.

    \highspace
    However, since $A$ is a quantum gate, it must be unitary (property of quantum gates):
    \begin{equation*}
        A^\dagger A = I
    \end{equation*}
    Assuming $A$ is real, we have $A^\dagger = A^T$, so:
    \begin{equation*}
        A^T =
        \begin{pmatrix}
            a_{11} & 0 \\
            a_{12} & a_{22}
        \end{pmatrix}
    \end{equation*}
    Hence:
    \begin{equation*}
        A^T A
        =
        \begin{pmatrix}
            a_{11} & 0 \\
            a_{12} & a_{22}
        \end{pmatrix}
        \cdot
        \begin{pmatrix}
            a_{11} & a_{12} \\
            0      & a_{22}
        \end{pmatrix}
        =
        \begin{pmatrix}
            a_{11}^{2} & a_{11}a_{12} \\
            a_{11}a_{12} & a_{12}^{2} + a_{22}^{2}
        \end{pmatrix}
    \end{equation*}
    Since $A$ is unitary (i.e., $A^\dagger A = I$), this matrix must equal the identity:
    \begin{equation*}
        \begin{pmatrix}
            a_{11}^{2} & a_{11}a_{12} \\
            a_{11}a_{12} & a_{12}^{2} + a_{22}^{2}
        \end{pmatrix}
        =
        \begin{pmatrix}
            1 & 0 \\
            0 & 1
        \end{pmatrix}
    \end{equation*}
    Therefore, to satisfy this equality, the following equations must hold:
    \begin{equation*}
        \begin{cases}
            a_{11}^{2} = 1 \\
            a_{11}a_{12} = 0 \\
            a_{12}^{2} + a_{22}^{2} = 1
        \end{cases}
    \end{equation*}
    From $a_{11}^{2}=1$, we get $a_{11}=\pm 1$, so $a_{11}\neq 0$. Therefore, from $a_{11}a_{12}=0$, it follows that:
    \begin{equation*}
        a_{12}=0
    \end{equation*}
    Substituting this into the last equation gives:
    \begin{equation*}
        a_{22}^{2}=1
        \quad \Rightarrow \quad
        a_{22}=\pm 1
    \end{equation*}
    Hence the matrix becomes diagonal:
    \begin{equation*}
        A =
        \begin{pmatrix}
            \pm 1 & 0 \\
            0 & \pm 1
        \end{pmatrix}
    \end{equation*}
    Since $A$ is diagonal, it only modifies the phases of $\ket{0}$ and $\ket{1}$, without mixing them. Thus, when we apply $A$ to $\ket{0}$ and $\ket{1}$, we get:
    \begin{equation*}
        A\ket{0} = \pm \ket{0},
        \qquad
        A\ket{1} = \pm \ket{1}
    \end{equation*}
    So neither $\ket{0}$ nor $\ket{1}$ is mapped to a superposition, because the output states are just $\ket{0}$ or $\ket{1}$ (up to a global phase). Thus, $A$ cannot create superposition (i.e., it cannot produce a state with both $\ket{0}$ and $\ket{1}$ components).
\end{proof}