\subsection{Querying Device Properties}

Different GPUs have varying capabilities and configurations, such as the number of Streaming Multiprocessors (SMs), maximum thread dimensions, memory size, etc. Knowing these properties helps developers optimize their code to run efficiently on different hardware. The main properties are:
\begin{itemize}
    \item \important{Get Device ID}: \texttt{cudaGetDevice}; to retrieve the current device ID.
    \begin{lstlisting}[language=C++]
int deviceId;
cudaGetDevice(&deviceId);
    \end{lstlisting}

    \item \important{Query Device Properties}: \texttt{cudaGetDeviceProperties}; to get the properties of the specified device.
    \begin{lstlisting}[language=C++]
cudaDeviceProp props;
cudaGetDeviceProperties(&props, deviceId);
    \end{lstlisting}

    \item \important{Understanding the \texttt{cudaDeviceProp} Structure}. This structure contains numerous attributes of the device. Some key attributes include:
    \begin{itemize}
        \item \texttt{name}. Name of the device.
        \item \texttt{totalGlobalMem}. Total global memory available on the device.
        \item \texttt{sharedMemPerBlock}. Shared memory available per block.
        \item \texttt{regsPerBlock}. Number of registers available per block.
        \item \texttt{warpSize}. Number of threads in a warp.
        \item \texttt{maxThreadsPerBlock}. Maximum number of threads per block.
        \item \texttt{maxThreadsDim}. Maximum dimension sizes of a thread block (x, y, z).
        \item \texttt{maxGridSize}. Maximum dimension sizes of a grid (x, y, z).
        \item \texttt{clockRate}. Clock frequency of the device.
        \item \texttt{multiProcessorCount}. Number of streaming multiprocessors on the device.
    \end{itemize}
\end{itemize}

\begin{examplebox}[: common queries]
    Some common queries are:
    \begin{itemize}
        \item Maximum Dimensions of a Thread Block:
        \begin{lstlisting}[language=C++]
printf("Max Thread Dimensions: %d x %d x %d\n",
        props.maxThreadsDim[0], props.maxThreadsDim[1],
        props.maxThreadsDim[2]);
        \end{lstlisting}

        \item Maximum Number of Threads in a Block:
        \begin{lstlisting}[language=C++]
printf("Max Threads per Block: %d\n", props.maxThreadsPerBlock);
        \end{lstlisting}

        \item Maximum Size of GPU Global Memory:
        \begin{lstlisting}[language=C++]
printf("Total Global Memory: %zu bytes\n", props.totalGlobalMem);
        \end{lstlisting}

        \item Warp Size:
        \begin{lstlisting}[language=C++]
printf("Warp Size: %d\n", props.warpSize);
        \end{lstlisting}

        \item Number of Streaming Multiprocessors (SM):
        \begin{lstlisting}[language=C++]
printf("Number of SMs: %d\n", props.multiProcessorCount);
        \end{lstlisting}

    \end{itemize}
\end{examplebox}
