\subsection{Parallel Patterns}

\begin{flushleft}
    \textcolor{Green3}{\faIcon{question-circle} \textbf{What are Parallel Patterns?}}
\end{flushleft}
\definition{Parallel Patterns} are essentially \textbf{recurring combinations of task distribution and data access strategies that can be applied to solve specific problems} in parallel computing. These patterns act as a \dquotes{vocabulary} for algorithm design, making it easier to communicate and understand various approaches to parallelism. 

\highspace
By applying these patterns, developers can achieve efficient and effective parallelism, ensuring that tasks are executed concurrently without conflicts and with optimal performance. 
\begin{itemize}[label=\textcolor{Green3}{\faIcon{check}}]
    \item \textcolor{Green3}{\textbf{Simplicity}}. They provide clear, reusable templates for designing parallel algorithms.
    \item \textcolor{Green3}{\textbf{Efficiency}}. They help optimize performance by utilizing parallelism effectively.
    \item \textcolor{Green3}{\textbf{Universality}}. They are applicable across different parallel programming systems and platforms.
\end{itemize}
In essence, \textbf{parallel patterns help streamline the process of designing and implementing parallel algorithms}, making it \hl{easier to manage dependencies and achieve concurrent execution}.

\highspace
Roughly speaking, parallel patterns are \textbf{similar to design patterns} in computer science, but they focus specifically on parallel computing. They provide reusable solutions to common problems in parallel programming, with the goal of simplifying and streamlining the development process.

\highspace
\begin{flushleft}
    \textcolor{Green3}{\faIcon{stream} \textbf{Types of Parallel Patterns}}
\end{flushleft}
\begin{enumerate}
    \item \important{Nesting Pattern}: Combining different parallel patterns within one another.
    \item \important{Serial/Parallel Control Patterns}: Handling control flow in parallel and serial region of code.
    \item \important{Serial/Parallel Data Management Patterns}: Managing data access and modifications in parallel and serial contexts.
    \item Other Patterns:
    \begin{multicols}{2}
        \begin{itemize}
            \item Superscalar Sequences
            \item Futures
            \item Speculative Selection
            \item Workpile
            \item Search
            \item Segmentation
            \item Expand
            \item Category Reduction
        \end{itemize}
    \end{multicols}
    \item \important{Programming Model Support for Patterns}: Tools and models that support implementing these patterns.
\end{enumerate}
